**Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces**

### 2025-06-12 analysis/2_study2.qmd
update

- p: poly1 male
- confright: poly2 male
- confleft: poly1 male (neg)
- bex: poly1 male
- Suggests that (related to the original FakeFace1 hypothesis) reality would be driven by extremes degrees of attractiveness.

@@

  p <- modelbased::estimate_prediction(m, data = newdata, predict = "mu", iterations = 50)
  confright <- modelbased::estimate_prediction(m, data = newdata, predict = "confright", iterations = 50)
  confleft <- modelbased::estimate_prediction(m, data = newdata, predict = "confleft", iterations = 50)
  precright <- modelbased::estimate_prediction(m, data = newdata, predict = "precright", iterations = 50)
  precleft <- modelbased::estimate_prediction(m, data = newdata, predict = "precleft", iterations = 50)
  if ("pex" %in% names(insight::find_parameters(m))) {
    pex <- modelbased::estimate_prediction(m, data = newdata, predict = "pex", iterations = 50)
  } else {
    pex <- list(Predicted = rep(0, nrow(newdata)))
  }
  if ("bex" %in% names(insight::find_parameters(m))) {
    bex <- modelbased::estimate_prediction(m, data = newdata, predict = "bex", iterations = 50)
  } else {
    bex <- list(Predicted = rep(0.5, nrow(newdata)))
  }
  if ("pmid" %in% names(insight::find_parameters(m))) {
    pmid <- modelbased::estimate_prediction(m, data = newdata, predict = "pmid", iterations = 50)
  } else {
    pmid <- list(Predicted = rep(0, nrow(newdata)))
  }

@@

pred <- visualize_dchoco(m1_chocoat,
                         newdata=insight::get_datagrid(m1_chocoat, length = 5, preserve_range = FALSE))

p3_density <- pred$densities |> 
  mutate(Attractive = as.factor(Attractive),
         y = ifelse(x %in% c(0, 1), y * 3, y)) |>  # Inflate extremes for visibility
  ggplot(aes(x=x, y=y)) +
  geom_line(aes(color=Attractive), linewidth = 1.25) +
  geom_path(data = pred$mu, aes(group = side), linewidth=0, lineend = 'round', 
            arrow = arrow(length = unit(0.025, "npc"), type = "closed")) +
  facet_wrap(~Sex) +
  labs(y = "Distribution", x = "Do you think this image is real?", color = "Attractiveness") +
  scale_colour_manual(values=scales::seq_gradient_pal("#FFCDD2", "#B71C1C", "Lab")(seq(0,1,length.out=5))) +
  scale_x_continuous(expand = c(0, 0), breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c("      0%", "25%", "50%", "75%", "100%      ")) +
  scale_linewidth_continuous(range = c(0.5, 2), guide = "none") +
  theme_minimal() +
  theme(axis.title.x = element_text(face = "bold"), 
        axis.text.y = element_blank(), 
        # axis.title.y = element_blank(), 
        strip.background = element_rect(fill = "lightgrey", color = NA)) 
p3_density
```

@@

colors <- c("p" = "#9C27B0", "confright" = "#795548", "confleft" = "#795548",
            "precright" = "#607D8B","precleft" = "#607D8B", "pex" = "#FF9800", "bex" = "#F44336", "pmid" = "black")

plot_params1 <- function(m, xlabel = "Attractiveness") {
  pred_params <- data.frame()
  for(param in c("mu", "confright", "confleft", "precright", "precleft", "pex", "bex")) {
    pred_params <- m |> 
      modelbased::estimate_prediction(data = "grid", length = 20, preserve_range = FALSE, predict = param, iterations = 10) |>
      as.data.frame() |> 
      dplyr::mutate(Parameter = param) |> 
      rbind(pred_params)
  }

@@

  pred_params |> 
    mutate(Parameter = ifelse(Parameter == "mu", "p", Parameter),
           Parameter = fct_relevel(Parameter, "p", "confright", "precright", "pex", "confleft", "precleft", "bex")) |> 
    filter(Parameter != "bex") |>
    ggplot(aes(x = x, y = Predicted)) +
    geom_ribbon(aes(ymin = CI_low, ymax = CI_high, fill = Parameter, group = Sex), alpha = 0.2) +
    geom_line(aes(color = Parameter, linetype = Sex), linewidth = 1) +
    facet_wrap(~Parameter, scales = "free_y", ncol=3, strip.position = "left") +
    labs(x = xlabel) +
    scale_fill_manual(values = colors, guide = "none") +
    scale_color_manual(values = colors, guide = "none") +
    # scale_x_continuous(expand = c(0, 0), breaks = c(0, 0.5, 1), labels = c("      0%", "50%", "100%      ")) +
    theme_minimal() +
    theme(strip.placement = "outside",
          axis.title.y = element_blank(),
          axis.title.x = element_text(face = "bold"),
          strip.text = element_text(size = rel(1), face = "italic"),
          panel.grid.minor.x = element_blank())
}

plot_params1(m1_chocoat, xlabel = "Attractiveness")
```

@@

plot_params1(m1_chocobeauty, xlabel = "Beauty")
```

@@

estimate_relation(m, length = 30) |> 
  ggplot(aes(x=Attractive, y = Predicted)) + 
  geom_ribbon(aes(ymin = CI_low, ymax = CI_high,, fill = Condition), alpha = 0.2) +
  geom_line(aes(color = Condition), linewidth = 1) +
  geom_smooth(data=dfraw, aes(color = Condition, y = Real), method = "loess", se = FALSE, linetype = "dotted", formula = 'y ~ x') +
  scale_color_manual(values = c("AI-Generated"="#3F51B5", "Photograph"="#FF5722")) +
  scale_fill_manual(values = c("AI-Generated"="#3F51B5", "Photograph"="#FF5722")) +
  facet_wrap(~Sex) +
  theme_minimal()
```

@@

```{r}
# m2_betagate <- readRDS("models/sample2_betagate.rds")
m2_betagate <- readRDS("models/sample2noex_betagate.rds")
m2_chocoat <- readRDS("models/sample2_chocoattractiveness.rds")
m2_chocoat <- readRDS("models/sample2noex_chocoattractiveness.rds")
```

@@

```{r}
c <- data.frame()
for(param in c("mu", "confright", "confleft", "precright", "precleft", "pmid")) {
  c <- estimate_contrasts(m2_chocoat, contrast = "Condition", by = "Sex", predict = param, backend = "emmeans") |>
    mutate(Parameter = ifelse(param == "mu", "p", param)) |> 
    rbind(c)
}
c
```

@@

```{r}
pred <- visualize_dchoco(m2_chocoat,
                         newdata=insight::get_datagrid(m2_chocoat, length = 5, preserve_range = FALSE),
                         x = seq(0, 1, length.out=9))

p4_density <- pred$densities |> 
  mutate(Attractive = as.factor(Attractive),
         y = ifelse(x %in% c(0, 1), y * 3, y),  # Inflate extremes for visibility
         x = ifelse(Condition == "AI-Generated", x + 0.01, x)) |>  
  filter(Condition == "Photograph") |>
  ggplot(aes(x=x, y=y)) +
  geom_line(aes(color=Attractive, linetype = Condition), linewidth = 1.25) +
  # geom_path(data = filter(pred$mu, Condition == "Photograph"), aes(group = side), linewidth=0, lineend = 'round', 
  #           arrow = arrow(length = unit(0.025, "npc"), type = "closed")) +
  facet_wrap(~Sex) +
  labs(y = "Distribution", x = "Do you think this image is real?", color = "Attractiveness") +
  scale_colour_manual(values=scales::seq_gradient_pal("#FFCDD2", "#B71C1C", "Lab")(seq(0,1,length.out=5))) +
  scale_x_continuous(expand = c(0, 0), breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c("      0%", "25%", "50%", "75%", "100%      ")) +
  scale_linewidth_continuous(range = c(0.5, 2), guide = "none") +
  theme_minimal() +
  theme(axis.title.x = element_text(face = "bold"), 
        axis.text.y = element_blank(), 
        # axis.title.y = element_blank(), 
        strip.background = element_rect(fill = "lightgrey", color = NA)) 
p4_density
```

@@

@@

```{r}
m <- lme4::lmer(Real ~ Sex / poly(Beauty, 2) * Condition + (1|Participant) + (1|Item), data = df)
display(parameters::parameters(m, effects = "fixed"))

@@

:::
## Figures

p3_a <- wrap_elements(p3_density / p3_params + 
         patchwork::plot_annotation(title = "Effect of Attractiveness", theme = theme(plot.title = element_text(face = "bold"))))
```

>>>>

@@

  p <- modelbased::estimate_prediction(m, data = newdata, predict = "mu", iterations = 100)
  confright <- modelbased::estimate_prediction(m, data = newdata, predict = "confright", iterations = 100)
  confleft <- modelbased::estimate_prediction(m, data = newdata, predict = "confleft", iterations = 100)
  precright <- modelbased::estimate_prediction(m, data = newdata, predict = "precright", iterations = 100)
  precleft <- modelbased::estimate_prediction(m, data = newdata, predict = "precleft", iterations = 100)
  if ("pex" %in% names(insight::find_parameters(m))) {
    pex <- modelbased::estimate_prediction(m, data = newdata, predict = "pex", iterations = 100)
  } else {
    pex <- list(Predicted = rep(0, nrow(newdata)))
  }
  if ("bex" %in% names(insight::find_parameters(m))) {
    bex <- modelbased::estimate_prediction(m, data = newdata, predict = "bex", iterations = 100)
  } else {
    bex <- list(Predicted = rep(0.5, nrow(newdata)))
  }
  if ("pmid" %in% names(insight::find_parameters(m))) {
    pmid <- modelbased::estimate_prediction(m, data = newdata, predict = "pmid", iterations = 100)
  } else {
    pmid <- list(Predicted = rep(0, nrow(newdata)))
  }

@@

pred1 <- visualize_dchoco(m1_chocoat,
                          newdata=insight::get_datagrid(m1_chocoat, length = 5, preserve_range = FALSE))

p3_density <- pred1$densities |> 
  mutate(Attractive = as.factor(Attractive),
         y = ifelse(x %in% c(0, 1), y * 3, y)) |>  # Inflate extremes for visibility
  ggplot(aes(x=x, y=y)) +
  geom_line(aes(color=Attractive), linewidth = 1.25) +
  geom_path(data = pred1$mu, aes(group = side), linewidth=0.3, lineend = 'round', 
            arrow = arrow(length = unit(0.03, "npc"), type = "closed")) +
  facet_wrap(~Sex) +
  labs(y = "Distribution", x = "Do you think this image is real?", color = "Attractiveness",
       title = "Effect on CHOCO distribution") +
  scale_colour_manual(values=scales::seq_gradient_pal("#FFCDD2", "#B71C1C", "Lab")(seq(0, 1, length.out=5))) +
  scale_x_continuous(expand = c(0, 0), breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c("      0%", "25%", "50%", "75%", "100%      ")) +
  # scale_linewidth_continuous(range = c(0.5, 2), guide = "none") +
  theme_minimal() +
  theme(axis.title.x = element_text(face = "bold"), 
        axis.text.y = element_blank(), 
        plot.title = element_text(face = "bold"),
        strip.background = element_rect(fill = "lightgrey", color = NA)) 
p3_density
```

@@

colors <- c("p" = "#9C27B0", "confright" = "#795548", "confleft" = "#795548",
            "precright" = "#607D8B","precleft" = "#607D8B", "pex" = "#FF9800", "bex" = "#FF7811", "pmid" = "#4CAF50")

plot_params1 <- function(m, xlabel = "Attractiveness") {
  pred_params <- data.frame()
  for(param in c("mu", "confright", "confleft", "precright", "precleft", "pex", "bex")) {
    pred_params <- m |> 
      modelbased::estimate_prediction(data = "grid", length = 40, preserve_range = FALSE, 
                                      predict = param, iterations = 200) |>
      as.data.frame() |> 
      dplyr::mutate(Parameter = param) |> 
      rbind(pred_params)
  }

@@

  pred_params |> 
    mutate(Parameter = ifelse(Parameter == "mu", "p", Parameter),
           Parameter = fct_relevel(Parameter, "p", "confright", "precright", "bex", "confleft", "precleft", "pex")) |> 
    filter(Parameter != "pex") |>
    ggplot(aes(x = x, y = Predicted)) +
    geom_ribbon(aes(ymin = CI_low, ymax = CI_high, fill = Parameter, group = Sex), alpha = 0.2) +
    geom_line(aes(color = Parameter, linetype = Sex), linewidth = 1) +
    facet_wrap(~Parameter, scales = "free_y", ncol=3, strip.position = "left") +
    labs(x = xlabel, title = "Effect on model parameters") +
    scale_fill_manual(values = colors, guide = "none") +
    scale_color_manual(values = colors, guide = "none") +
    scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c(" 0", "0.25", "0.5", "0.75", "1 ")) +
    theme_minimal() +
    theme(strip.placement = "outside",
          axis.title.y = element_blank(),
          axis.title.x = element_text(face = "bold"),
          plot.title = element_text(face = "bold"),
          strip.text = element_text(size = rel(1), face = "italic"),
          panel.grid.minor.x = element_blank())
}

p3_params <- plot_params1(m1_chocoat, xlabel = "Attractiveness")

# Add significance
sig <- data.frame(
  Sex = c("Male", "Male", "Male", "Male", "Male"),
  Label = c("/", "U", "\\", "∩", "/"),
  Parameter = c("p", "confright", "confleft", "precleft", "bex"),
  x = 0.5,
  y = c(0.51, 0.6, 0.63, 3.75, 0.42)
) |> 
  mutate(Parameter =  fct_relevel(Parameter, "p", "confright", "bex", "confleft", "precleft"))

p3_paramssig <- p3_params + 
  geom_text(data = sig, aes(x = x, y = y, label = Label), 
            size = 5, color = "black", hjust = 0.5) 
p3_paramssig
```

@@

@@

estimate_relation(m, length = 30) |> 
  ggplot(aes(x=Attractive, y = Predicted)) + 
  geom_ribbon(aes(ymin = CI_low, ymax = CI_high,, fill = Condition), alpha = 0.2) +
  geom_line(aes(color = Condition), linewidth = 1) +
  geom_smooth(data=df, aes(color = Condition, y = RealNoex), method = "loess", se = FALSE, linetype = "dotted", formula = 'y ~ x') +
  scale_color_manual(values = c("AI-Generated"="#3F51B5", "Photograph"="#FF5722")) +
  scale_fill_manual(values = c("AI-Generated"="#3F51B5", "Photograph"="#FF5722")) +
  facet_wrap(~Sex) +
  theme_minimal()
```

@@

Interestingly, Beta-gate appears as a better fit, but this is likely related to the nature of the index used, WAIC, which heavily penalizes for the number of parameters (which is *much* larger for CHOCO). Probably not a very meaningful result, especially since the posterior checks clearly show its misfit.

```{r}
# m2_betagate <- readRDS("models/sample2_betagate.rds")
m2_betagate <- readRDS("models/sample2noex_betagate.rds")
# m2_chocoat <- readRDS("models/sample2_chocoattractiveness.rds")
m2_chocoat <- readRDS("models/sample2noex_chocoattractiveness.rds")
```

@@

```{r}
c <- data.frame()
for(param in c("mu", "confright", "confleft", "precright", "precleft", "pmid")) {
  c <- estimate_contrasts(m2_chocoat, contrast = "Condition", by = "Sex", predict = param, 
                          backend = "emmeans", test = "pd") |>
    mutate(Parameter = ifelse(param == "mu", "p", param)) |> 
    rbind(c)
}
c
```

```{r}
estimate_contrasts(m2_chocoat, contrast = "Condition", by = c("Attractive", "Sex"), 
                   predict = "mu", length = 5, backend = "emmeans", test = "pd") 
```

@@

```{r}
pred2 <- visualize_dchoco(m2_chocoat,
                          newdata=insight::get_datagrid(m2_chocoat, length = 5, preserve_range = FALSE))

p4_density <- pred2$densities |> 
  mutate(Attractive = as.factor(Attractive)) |> 
         # y = ifelse(x %in% c(0, 1), y * 3, y),  # Inflate extremes for visibility
         # x = ifelse(Condition == "AI-Generated", x + 0.01, x)) |>  
  # filter(Condition == "Photograph") |>
  ggplot(aes(x=x, y=y)) +
  geom_line(aes(color=Attractive, linetype = Condition, linewidth = Condition)) +
  # geom_path(data = filter(pred2$mu, Condition == "Photograph"), aes(group = side), linewidth=0, lineend = 'round', 
  #           arrow = arrow(length = unit(0.025, "npc"), type = "closed")) +
  facet_wrap(~Sex) +
  labs(y = "Distribution", x = "Do you think this image is real?", color = "Attractiveness",
       title = "Effect on CHOCO distribution") +
  scale_colour_manual(values=scales::seq_gradient_pal("#FFCDD2", "#B71C1C", "Lab")(seq(0,1,length.out=5))) +
  scale_x_continuous(expand = c(0, 0), breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c("      0%", "25%", "50%", "75%", "100%      ")) +
  scale_linetype_manual(values = c("Photograph" = "solid", "AI-Generated" = "dotted")) +
  scale_linewidth_manual(values = c("Photograph" = 1.25, "AI-Generated" = 0.8)) +
  theme_minimal() +
  theme(axis.title.x = element_text(face = "bold"), 
        axis.text.y = element_blank(),  
        plot.title = element_text(face = "bold"),
        # axis.title.y = element_blank(), 
        strip.background = element_rect(fill = "lightgrey", color = NA)) 
p4_density
```

@@

plot_params2 <- function(m, xlabel = "Attractiveness") {
  pred_params <- data.frame()
  for(param in c("mu", "confright", "confleft", "precright", "precleft", "pmid")) {
    pred_params <- m |> 
      modelbased::estimate_prediction(data = "grid", length = 40, preserve_range = FALSE, predict = param, iterations = 200) |>
      as.data.frame() |> 
      dplyr::mutate(Parameter = param) |> 
      rbind(pred_params)
  }

  pred_params |> 
    mutate(Parameter = ifelse(Parameter == "mu", "p", Parameter),
           Parameter = fct_relevel(Parameter, "p", "confright", "precright", "pmid", "confleft", "precleft")) |> 
    ggplot(aes(x = x, y = Predicted)) +
    geom_ribbon(aes(ymin = CI_low, ymax = CI_high, fill = Parameter, 
                    group = interaction(Sex, Condition), alpha = Condition)) +
    geom_line(aes(color = Parameter, linetype = Sex, 
                  alpha = interaction(Sex, Condition), group = interaction(Sex, Condition)), linewidth = 1) +
    facet_wrap(~Parameter, scales = "free_y", ncol=3, strip.position = "left") +
    labs(x = xlabel, title = "Effect on model parameters") +
    scale_fill_manual(values = colors, guide = "none") +
    scale_color_manual(values = colors, guide = "none") +
    scale_alpha_manual(values = c("Photograph" = 0.2, "AI-Generated" = 0, 
                                  "Male.Photograph" = 1, "Female.Photograph" = 1,
                                  "Male.AI-Generated" = 0.2, "Female.AI-Generated"=0.2), guide = "none") +
    scale_x_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c(" 0", "0.25", "0.5", "0.75", "1 ")) +
    theme_minimal() +
    theme(strip.placement = "outside",
          axis.title.y = element_blank(),
          axis.title.x = element_text(face = "bold"),
          plot.title = element_text(face = "bold"),
          strip.text = element_text(size = rel(1), face = "italic"),
          panel.grid.minor.x = element_blank())
}

p4_params <- plot_params2(m2_chocoat, xlabel = "Attractiveness")

# Add significance
sig <- data.frame(
  Sex = c("Male", "Male", "Female", "Male"),
  Label = c("/", "/", "U", "\\"),
  Parameter = c("p", "confright", "confright", "confleft"),
  x = 0.5,
  y = c(0.48, 0.425, 0.49, 0.48)
) |> 
  mutate(Parameter =  fct_relevel(Parameter, "p", "confright", "confleft"))

p4_paramssig <- p4_params + 
  geom_text(data = sig, aes(x = x, y = y, label = Label), 
            size = 5, color = "black", hjust = 0.5) 
p4_paramssig
```

@@

```{r}
m <- lme4::lmer(RealNoex ~ Sex / poly(Beauty, 2) * Condition + (1|Participant) + (1|Item), data = df)
display(parameters::parameters(m, effects = "fixed"))

@@

```{r}
table_dvour(modelbased::estimate_grouplevel(m2_chocobeauty))
```

```{r}
summary(m2_chocobeauty, robust = TRUE)
```

```{r}
pred <- visualize_dchoco(m2_chocobeauty,
                         newdata=insight::get_datagrid(m2_chocobeauty, length = 5, preserve_range = FALSE))

p5_density <- pred$densities |> 
  mutate(Beauty = as.factor(Beauty)) |> 
         # y = ifelse(x %in% c(0, 1), y * 3, y),  # Inflate extremes for visibility
         # x = ifelse(Condition == "AI-Generated", x + 0.01, x)) |>  
  # filter(Condition == "Photograph") |>
  ggplot(aes(x=x, y=y)) +
  geom_line(aes(color=Beauty, linetype = Condition, linewidth = Condition)) +
  # geom_path(data = filter(pred$mu, Condition == "Photograph"), aes(group = side), linewidth=0, lineend = 'round', 
  #           arrow = arrow(length = unit(0.025, "npc"), type = "closed")) +
  facet_wrap(~Sex) +
  labs(y = "Distribution", x = "Do you think this image is real?", color = "Attractiveness",
       title = "Effect on CHOCO distribution") +
  scale_colour_manual(values=scales::seq_gradient_pal("#C5CAE9", "#283593", "Lab")(seq(0,1,length.out=5))) +
  scale_x_continuous(expand = c(0, 0), breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c("      0%", "25%", "50%", "75%", "100%      ")) +
  scale_linetype_manual(values = c("Photograph" = "solid", "AI-Generated" = "dotted")) +
  scale_linewidth_manual(values = c("Photograph" = 1.25, "AI-Generated" = 0.8)) +
  theme_minimal() +
  theme(axis.title.x = element_text(face = "bold"), 
        axis.text.y = element_blank(),  
        plot.title = element_text(face = "bold"),
        # axis.title.y = element_blank(), 
        strip.background = element_rect(fill = "lightgrey", color = NA)) 
p5_density
```

## Figures

```{r}
#| eval: true
#| cache-lazy: false

p3_a <- wrap_elements((p3_density + theme(legend.position = "none")) / 
                        (p3_paramssig + theme(legend.position = "none")) + 
         patchwork::plot_annotation(title = "Sample 1", 
                                    theme = theme(plot.title = element_text(face = "bold.italic", hjust = 0.5))))

p4_a <- wrap_elements((p4_density + theme(axis.title.y = element_blank())) / 
                        p4_paramssig + 
         patchwork::plot_annotation(title = "Sample 2", 
                                    theme = theme(plot.title = element_text(face = "bold.italic", hjust = 0.5))))

p5 <- (p3_a | p4_a) + 
  plot_layout(guides = "collect", widths = c(0.45, 0.55))

ggsave("../paper/figures/fig5.png", p5, width = 13, height = 11, dpi = 300)
```

### 2025-06-12 paper/manuscript.qmd
update

---
title: "**Introducing the Choice-Confidence (CHOCO) Model for Bimodal Scales Data: Application to the Relationship between Reality Beliefs about AI-Generated Faces and Beauty**"
shorttitle: "Choice-Confidence (CHOCO) Model"
author:
  - name: Dominique Makowski
    corresponding: true
    orcid: 0000-0001-5375-9967
    email: D.Makowski@sussex.ac.uk
    url: https://realitybending.github.io/
    roles:
      - Conceptualization
      - Data curation
      - Formal Analysis
      - Funding acquisition
      - Investigation
      - Methodology
      - Project administration
      - Resources
      - Software
      - Supervision
      - Validation
      - Visualization
      - Writing – original draft
    affiliations:
      - id: "id1"
        name: "University of Sussex"
        department: "School of Psychology"
        city: "Brighton"
        country: "UK"
        postal-code: "BN1 9RH"
      - name: "University of Sussex"
        department:  "Sussex Centre for Consciousness Science"
  - name: Ana Neves  
    orcid: 0009-0006-0020-7599
    role:
      - Data curation
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
  - name: Andy Field 
    orcid: ""
    role:
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    gratitude: |
      ::: {.callout-note icon=false appearance="simple"}
      This preprint is a non-peer-reviewed work from the [**Reality Bending Lab**](https://realitybending.github.io/).
      ![](https://realitybending.github.io/media/ReBeL_LogoOnly_hu11484441381606756729.png){width=20% fig-align="center"}
      :::
    study-registration: ~
    data-sharing: ~
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    authorship-agreements: ~
abstract: |
  TO DO.
#   **Significance Statement**. TODO.
keywords: "subjective scales, choice confidence model, reality beliefs, AI-generated faces, attractiveness"
numbered-lines: true
floatsintext: true
mask: false
bibliography: bibliography.bib
format:
  apaquarto-pdf:
    documentmode: jou
    keep-tex: true
  apaquarto-docx: default
---

@@

<!-- Problem statement -->
Despite significant advancements in psychological science following the replication crisis [@open2015estimating], its progress is still hindered by its sub-optimal (or inappropriate) usage of statistical tools [@makowski2023we]. 
A prevalent issue is the continued reliance on linear models that assume normally distributed (Gaussian) data^[More specifically, that the outcome is distributed according to a Normal distribution which parameters are expressed as a linear function of the predictors.] - as this assumption often does not hold true for many types of psychological outcomes.
For instance, reaction times typically exhibit skewed distributions, choices can be represented as binary variables, and count data consists of strictly positive integers. 
Applying models that presume normality and model the "mean" of the outcome variable can lead to misinterpretation and potentially misleading conclusions when applied indiscriminately. It is thus important that psychologists use models that can best describe (or generate) the data they collect, to fully exploit them and bring more nuance and accuracy to their conclusions.

@@

Using the open-accss data from @makowski2025too, we included all heterosexual and bisexual (as these two groups did not seem to differ based on preliminary analyses and were thus grouped to maximize power) male and female participants, for a final sample of 141 participants (Mean age = 28.4, SD = 9.0, range: [19, 66]; Sex: 47.5% females). For each participant, we included only stimuli of the opposite gender (i.e., all 89 female faces for men and 20 male faces for women).

@@

@@

>>>>

---
title: "**Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces**"
shorttitle: "Choice-Confidence (CHOCO) Model"
author:
  - name: Dominique Makowski
    corresponding: true
    orcid: 0000-0001-5375-9967
    email: D.Makowski@sussex.ac.uk
    url: https://realitybending.github.io/
    roles:
      - Conceptualization
      - Data curation
      - Formal Analysis
      - Funding acquisition
      - Investigation
      - Methodology
      - Project administration
      - Resources
      - Software
      - Supervision
      - Validation
      - Visualization
      - Writing – original draft
    affiliations:
      - id: "id1"
        name: "University of Sussex"
        department: "School of Psychology"
        city: "Brighton"
        country: "UK"
        postal-code: "BN1 9RH"
      - name: "University of Sussex"
        department:  "Sussex Centre for Consciousness Science"
  - name: Ana Neves  
    orcid: 0009-0006-0020-7599
    role:
      - Data curation
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
  - name: Andy Field 
    orcid: ""
    role:
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    gratitude: |
      ::: {.callout-note icon=false appearance="simple"}
      This preprint is a non-peer-reviewed work from the [**Reality Bending Lab**](https://realitybending.github.io/).
      ![](https://realitybending.github.io/media/ReBeL_LogoOnly_hu11484441381606756729.png){width=20% fig-align="center"}
      :::
    study-registration: ~
    data-sharing: ~
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    authorship-agreements: ~
abstract: |
  TO DO.
#   **Significance Statement**. TODO.
keywords: "subjective scales, choice confidence model, reality beliefs, AI-generated faces, attractiveness"
# Add more keywords if possible: beauty, Ordered Beta model, Bimodal data, decision-making
numbered-lines: true
floatsintext: true
mask: false
bibliography: bibliography.bib
format:
  apaquarto-pdf:
    documentmode: jou
    keep-tex: true
  apaquarto-docx: default
---

@@

<!-- Problem statement -->
Despite significant advancements in psychological science following the replication crisis [@open2015estimating], its progress is still hindered by its sub-optimal (or inappropriate) usage of statistical tools [@makowski2023we; @cumming2014new; @blanca2018current]. 
A prevalent issue is the continued reliance on linear models that assume normally distributed (Gaussian) data^[More specifically, that the outcome is distributed according to a Normal distribution which parameters are expressed as a linear function of the predictors.] - as this assumption often does not hold true for many types of psychological outcomes.
For instance, reaction times typically exhibit skewed distributions, choices can be represented as binary variables, and count data consists of strictly positive integers. 
Applying models that presume normality and model the "mean" of the outcome variable can lead to misinterpretation and potentially misleading conclusions when applied indiscriminately. It is thus important that psychologists use models that can best describe (or generate) the data they collect, to fully exploit them and bring more nuance and accuracy to their conclusions.

@@

Using the open-access data from @makowski2025too, we included all heterosexual and bisexual (as these two groups did not seem to differ based on preliminary analyses and were thus grouped to maximize power) male and female participants, for a final sample of 141 participants (Mean age = 28.4, SD = 9.0, range: [19, 66]; Sex: 47.5% females). For each participant, we included only stimuli of the opposite gender (i.e., all 89 female faces for men and 20 male faces for women).

@@

The findings of Study 2 replicate and extend those of Study 1, supporting the idea that facial attractiveness significantly shapes participants' beliefs about reality beliefs pertaining to human faces. This effect was particularly strong and robust among male participants. For females, the effect was weaker in Study 1 but more pronounced in Study 2, which might partly reflect increased statistical power.

Another possibility, however, is that the second sample's experimental manipulation introduced conflicting cues (e.g., faces labelled "AI-generated") that could have disrupted intuitive heuristics, thereby increasing reliance on attractiveness as a diagnostic cue, particularly in ambiguous trials.
<!-- This could be tested by analyzing believed vs. non-believed items -->

@@

Also: https://www.cell.com/neuron/fulltext/S0896-6273(16)30016-2
